---
title: "The BOD-O2 model"
subtitle: "A template R Markdown file for a simple dynamic model in 0-D"
author: "your name here"
date: "Date of creation here"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This template file contains a simple box model describing the dynamics of molecular oxygen (O$_2$, in $mol~m^{-3}$) and biochemical oxygen demand (BOD, in $mol~m^{-3}$) in a lake. It is assumed that 

* O$_2$ and BOD are removed due to BOD decay (O$_2$:BOD stoichiometry of 1:1),
* BOD decay is a first-order process with respect to BOD ($k_{Decay} = 0.05~d^{-1}$) and depends on O$_2$ according to the Michaelis-Menten kinetics ($K_{O2}=0.001~mol~m^{-3}$),
* O$_2$ is added by air-water exchange ($k_{Aeration}=0.1~d^{-1}$, $O_{2,sat} = 0.3~mol~m^{-3}$),
* BOD is added at a constant rate ($0.001~mol~m^{-3}~d^{-1}$),
* initial concentrations are $O_{2,ini}=0.25~mol~m^{-3}$ and $BOD_{ini} = 0.5~mol~m^{-3}$.

# Model definition

```{r, eval = TRUE, cache = FALSE, echo = TRUE, warning=FALSE, message=FALSE}
# units: time=days, amount=moles, concentration=mol/m3

F12 <- 70 
F21  <-  70.6
F12W <-  0.2 
F23  <-  50 
F32  <- 39
F34  <- 11
F24  <-  90.2
F42  <-  101
F45  <-  0.2
F56  <-  0.2
F62  <-  0.2
F72  <- 0.4
F71  <-  59.6
F17  <-  60

# Initial conditions of the state variables
state.ini <- c(M1.Atm <-  590, M2.SurfOcean <-  900, M3.Marinbiota <-  3, M4.Intdeepocean <-  37100, M5.Surfsediment <-  150, M6.SedRockOrg <-  75011000, M7.VegSoil <-  2300) # all in [PgC]

# Model parameters. all in [PgC/Year]
pars <- c(
  r12 = F12/M1.Atm, r12w = F12W/M1.Atm, r21 = F21/M2.SurfOcean, r23 = F23/M2.SurfOcean, r32 = F32/M3.Marinbiota, r34 = F34/M3.Marinbiota, r24 = F24/M2.SurfOcean, r42 = F42/M4.Intdeepocean, r45 = F45/M4.Intdeepocean, r56 = F56/M5.Surfsediment, r62 = F62/M6.SedRockOrg, r72 = F72/M7.VegSoil, r71 = F71/M7.VegSoil, r17 = F17/M1.Atm)

# Model function: calculates time-derivatives and other output
Ccyclemodel <-function(t, state, parms) { 
  # t: time, state: state variables at time t, parms: model parameters
  with (as.list(c(state, parms)),{
 
    # rate expressions [PgC/year]
    #Decay      <- kDecay * BOD * O2/(O2+K.O2) # BOD decay
    #Aeration   <- kAeration * (satO2-O2)      # air-water exchange

    # Time-derivatives: dC/dt = production - consumption [mol/m3/d]
    dM1.ATM.dt <- r21*M2.SurfOcean + r71*M7.VegSoil - r12w*M1.Atm - r17*M1.Atm - r12*M1.Atm
    
    dM2.SurfOcean.dt <- r12*M1.Atm + r32*M3.Marinbiota + r62*M6.SedRockOrg + r72*M7.VegSoil+ r12w*M1.Atm + r42*M4.Intdeepocean - r21*M2.SurfOcean - r23*M2.SurfOcean - r24*M2.SurfOcean
    
    dM3.Marinbiota.dt <- r23*M2.SurfOcean - r32*M3.Marinbiota - r34*M3.Marinbiota
    
    dM4.Intdeepocean.dt <- r24*M2.SurfOcean + r34*M3.Marinbiota - r42*M4.Intdeepocean - r45*M4.Intdeepocean
    
    dM5.Surfsediment.dt <-  r45*M4.Intdeepocean -r56*M5.Surfsediment
    
    dM6.SedRockOrg.dt <-  r56*M5.Surfsediment - r62*M6.SedRockOrg
    
    dM7.VegSoil.dt <- r17*M1.Atm - r71*M7.VegSoil - r72*M7.VegSoil

    # return time-derivatives and ordinary variables as a list     
    return(list(c(dM1.ATM.dt, dM2.SurfOcean.dt, dM3.Marinbiota.dt, dM4.Intdeepocean.dt, dM5.Surfsediment.dt, dM6.SedRockOrg.dt, dM7.VegSoil.dt), # vector with derivatives 
                                    # (the same order as state variables!)
      # other output
      Sum = M1.Atm + M2.SurfOcean + M3.Marinbiota + M4.Intdeepocean + M5.Surfsediment + M6.SedRockOrg + M7.VegSoil))    
  })
}
```

# Model solution

## Dynamic solution

We run the model dynamically over 1000 years, using two different values of the reaeration rate constant: 

```{r, message=FALSE}
require(deSolve)  # package with integration methods

outtimes <- seq(from = 1, to = 1000, length.out = 10) # vector of output times

# ode integrates the model
out <- ode(y = state.ini, parms = pars, func = Ccyclemodel, times = outtimes)                

# change the value of the reaeration rate constant
#pars2      <- pars                          # copy the original parameter vector
#pars2["kAeration"] <- pars2["kAeration"]/10 # 10-fold lower kAeration

# integrate the model with the new parameters
#out2 <- ode(y = state.ini, parms = pars2, func = BODmodel, times = outtimes)                
```

```{r, eval=FALSE}
# print summary of the solution
#summary(out)
```

We plot both solutions in one graph:

```{r, fig.height = 6, fig.width = 6, fig.align='center'}
plot(out, lwd=2, mfrow=c(2,4))  # plot solutions
#plot(out, out2, xlab="time (days)", las=1, lwd=2,
    # ylab=list("molO2/m3","molBOD/m3","molBOD/m3/d","molO2/m3/d"))
#legend("topright", legend = c("original", "10-fold lower"),
      # title="aeration rate constant:", 
       #col=1:2, lwd=2, lty=1:2, bty="n")
```

## Steady-state solution

We find the steady-state solution:

```{r, message=FALSE}
#require(rootSolve) # package with solution methods
#std <- steady(y = state.ini, parms = pars, func = BODmodel, 
 #             positive = TRUE) # to ensure that the solution is positive
#std$y
```

Sometimes, it may happen that there are *multiple* steady state solutions to a model (e.g., one with non-zero values, another one with zero values). In this case, the ``steady`` function is unable to "decide" which one to choose and stops with an error. To overcome this error, use the function ``runsteady`` instead to find the steady state solution.

```{r, message=FALSE}
#std <- runsteady(y = state.ini, parms = pars, func = BODmodel)
#std$y
```

## Sensitivity analysis

We perform a sensitivity analysis to find how the steady-state depends on the reaeration rate constant:

```{r}
k.seq   <- seq(from = 0.01, to = 0.1, length.out = 100)
BOD.seq <- vector()  # will contain the results
O2.seq  <- vector()

for (i in 1:length(k.seq)){
  # parameter values for this run
  p <- pars
  p["kAeration"] <- k.seq[i]  # reaeration rate constant based on the sequence
  # steady-state with new parameter values
  std <- steady(y = state.ini, parms = p, func = BODmodel, positive = TRUE)  
  BOD.seq[i] <- std$y["BOD"]
  O2.seq[i]  <- std$y["O2"]
}
```

Finally, we plot the results of the sensitivity analysis:

```{r, fig.width=7, fig.height=3.5}
par(mfrow=c(1,2)) # figures in 1 row, 2 columns
plot(k.seq, O2.seq,  type="l", xlab="kAeration [/d]", main="O2",  ylab="molO2/m3")
plot(k.seq, BOD.seq, type="l", xlab="kAeration [/d]", main="BOD", ylab="molBOD/m3")
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. https://www.R-project.org/.
  
Soetaert Karline (2009).  rootSolve: Nonlinear root finding, equilibrium and steady-state analysis of ordinary differential equations.  R-package version 1.6
  
Soetaert Karline, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25. http://www.jstatsoft.org/v33/i09/ DOI: 10.18637/jss.v033.i09
  
